\section{Conclusion}
Our research aims to identify potential improvements in the population size based EU vaccination policy against the COVID-19 pandemic such that deaths across countries can be minimized. We propose a deterministic SIRD compartment model with two countries and two vaccines to examine the effect of optimal strategies compared to the current EU strategy. We calibrate our model with parameters from the literature and the real-world vaccine inflow of the COVID-19 vaccines. We construct the vaccination channels via piecewise constant functions and splines. We examine one case where we impose further restrictions on the optimal vaccine allocation and another case where we constrain the results to be an improvement for each country. We simulate the models numerically and validate the optimal strategies derived from the deterministic model in a stochastic extension to our model.\\

Our results show that the optimal derived vaccination strategies differ from the current EU strategy, which could indicate that more complex vaccination policies can lower the number of death cases caused by the pandemic. Enhancing the robustness of our results, we qualitatively find highly similar results using piecewise constant and spline vaccination channels. Leaving aside country-specific interests, it appears that assigning most vaccine doses to one country can be as much as 17\% beneficial, leading to a reduction of approximately 80\% fewer deaths within this country. The other country then experiences a substantial increase in deaths of about 23\%. But since the decrease in the number of cases in the first country is substantially lower, this leads to an overall decrease in the number of deaths. However, policymakers of the second country would not be willing to agree to this policy due to the higher case numbers. We find that imposing additional Pareto constraints yield an overall improvement by 5\% in comparison to the current strategy but an overall deterioration in comparison to the unrestricted optimal strategy. The Pareto optimal strategy derived by the model assigns just enough vaccine doses to the second country such that it has the same case numbers as for the baseline EU strategy. However, the first country is better off, leading to an overall improvement of the Pareto strategy in comparison to the current strategy. Since both countries are not worse off, they have no incentive to vote against the Pareto optimal strategy, making it more likely to be implemented in practice. Stochastic simulations yielded higher death numbers but underlined the general findings with respect to the efficiency of the strategies.

In ongoing work, we pursue two avenues for further improvement. First, we implement neural networks as the third channel of vaccine allocation. The piecewise constant approach and the splines do not yield constructive vaccination strategies since we do not link the vaccine allocation directly to the states of the model. As inputs for the neural network model, we use states and parameters that could be (at least approximately) observed in the real world, such as the current infected case numbers, the change in infected case numbers, and the reproduction number. Due to the linkage of vaccine allocation output to the model states, an optimal allocation in this scenario is transferable to the real world in the sense that the real-world vaccination strategy can be adjusted according to the output of the neural network, given the real-world's state. Second, we plan to incorporate further compartments that account for different age structures of countries. Adding further age-dependent compartments may help us to understand how vaccination strategies must be adapted according to country-specific demographics. We especially aim to focus on the implications of child vaccinations.  \\

%Third, we aim to calibrate our model with data to derive the parameters empirically and to validate our simulations. In addition, we implement several 
%sensitivity checks, such as changing the sensitivity checks for several parameters: hyperbolic tangent instead of logistic function for splines, parameter values, distance!, multi-country example
%\textbf{What is currently in planning}
%\begin{itemize}

%\item stochastic optimization
%\end{itemize}


In addition, it might be worth examining how the dynamics of the model change after including testing and quarantines. We are particularly interested in examining how testing and vaccinations can be optimally exploited together or whether tests can even substitute for vaccinations up to a certain degree. Moreover, our analysis does not account for non-death related disutilities, like physical long-term damage caused by infection. A modified objective that takes long-term measures into account could help to address this issue. \\




Our findings regarding the unrestricted optimization trace out an important dilemma for policy makers of supranational institutions, such as the European Union, when it comes to choosing a vaccination strategy for the current COVID-19 pandemic. On the one hand, choosing the overall number of death cases as objective measure seems to be a plausible objective. On the other hand, the supranational policymakers cannot outweigh the disutility of one country with the benefits of another country. However, our findings concerning the Pareto optimality indicate that given the current strategy, a Pareto improvement might be possible to achieve and for policymakers to apply, if further research is conducted.


